Towards Verifying the Bitcoin-S Library

### 2020-07-24 paper-short/main.tex
work on final version

\relatedversion{A full version of the paper is available at
  \url{https://github.com/kaibr/bitcoin-s-verification/blob/master/paper/paper.pdf}.}

@@

%\funding{(Optional) general funding statement
% \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

@@

%\hideOASIcs  %uncomment to remove references to OASIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The example in Figure~\ref{fig:factorial} adapted from the Stainless
documentation \cite{Stainless:documentation} shows how the verifier is
used. Note how a precondition is specified using \texttt{require} and
a postcondition using \texttt{ensuring}. Our function does not satisfy
the specification. An overflow in the 32-bit integer type leads to a
negative result for the input 17, as Stainless reports in
Figure~\ref{fig:failed}. Changing the type \texttt{Int} to
\texttt{BigInt} will result in a successful verification.

@@

\paragraph{The Pure Scala Fragment.} The Scala fragment supported by
Stainless comprises algebraic data types in the form of abstract
classes, case classes and case objects, objects for grouping classes
and functions, boolean expressions with short-circuit interpretation,
generics with invariant type parameters, pattern matching, local and
anonymous classes and more.  In addition to Pure Scala Stainless also
supports some imperative features, such as using a (mutable) variable
in a local scope of a function and while loops. They turn out not to
be relevant for our current work.

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
can not generate new coins. We call it the \emph{no-inflation
  property}. Trying to verify it, we uncover and fix a bug in the
Bitcoin-S library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. So we turn to a simpler property to verify. The
simplest possible property we can think of is the fact that adding
zero satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{addition-with-zero property} and we try
to verify it in Section~3. Here as well we see that a significant part
of the code lies outside of the supported fragment. We rewrite it
until we arrive at code that we successfully verify. In that process
we find and fix a second bug in Bitcoin-S.

@@

The \texttt{checkTransaction} function shown in
Figure~\ref{fig:checktrans} is crucial for the verification of the
no-inflation property. Given a transaction it returns true if some
basic checks succeed, otherwise false. For example, one of those
checks is that both the list of inputs and list of outputs need to be
non-empty.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 15-17
in Figure~\ref{fig:checktrans}. Here, the value
\texttt{prevOutputTxIds} gathers a list of all transaction identifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict:
\texttt{checkTransaction} returns false for valid transactions.

@@

Specifically, we replace lines 15-17 as follows:
\begin{lstlisting}[style=scala, firstnumber=15]
  val prevOutputs = transaction.inputs.map(_.previousOutput)
  val noDuplicateInputs = 
    prevOutputs.distinct.size == prevOutputs.size
\end{lstlisting}

We submitted this fix together with a corresponding unit test to the
Bitcoin-S project in a pull request, which has been merged
\cite{BitcoinS:pull435}.

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire Bitcoin-S codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
no-inflation-property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment. We tried to transform the code into
the supported fragment, but quickly realized that a better approach is
to first verify a simpler property depending on less code and later
come back to the no-inflation property with more experience. So we now
turn to the addition-with-zero property.

\section{The Addition-with-Zero Property}

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in Bitcoin-S is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is the only relevant notion of correctness. The most widely
used bitcoin implementation by far is the reference implementation
Bitcoin Core, written in C++. For consensus code, Bitcoin-S has thus
little choice but to be in strict agreement with the reference
implementation. To achieve that, it implements C-like data types in
Scala and then implements functionality using those C-like data
types. For example, the Satoshis class, which represents an amount of
satoshis, is implemented using the class \texttt{Int64} which aims to
represent the C-type \texttt{int64\_t}.

@@

While this is a potentially serious bug, it turns out that
\texttt{checkResult} is only ever called inside a constructor call for
a number type which contains the intended range check, see lines
32-35. The \texttt{checkResult} function thus can, and should, be
removed entirely. The Bitcoin-S developers have acknowledged the bug
and we submitted a pull request to fix it \cite{BitcoinS:pull565}.

% For further development, this raises the question: if the goal of the
% \texttt{Int64} type is to emulate \texttt{int64\_t} then why does it
% prevent overflows? To achieve strict agreement with Bitcoin Core, a
% better approach might be to remove overflow checking from the data
% type and to add it in exactly those places where it happens in Bitcoin
% Core.

@@

All transformations are \emph{equivalent} in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation.

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those objects into
case objects. That transformation is equivalent: case objects have
various additional properties (for example, being serializable) but
none of our code depends on the absence of those.

@@

\paragraph{The Specification.} Now that all our code has been
transformed into the supported fragment, we can finally write our
specification, shown in Figure~\ref{fig:spec}, and verify it with
Stainless, as the output in Figure~\ref{fig:result} shows.

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{result_output}
	\caption{Stainless output for the transformed code}
  \label{fig:result}
\end{figure}

>>>>

% \relatedversion{A full version of the paper is available at
%   \url{https://github.com/kaibr/bitcoin-s-verification/blob/master/paper/paper.pdf}.}

@@

% \funding{(Optional) general funding statement
% \dots}%optional, to capture a funding statement, which applies to all authors.
%Please enter author specific funding statements as fifth argument of the \author macro.

@@

% \hideOASIcs %uncomment to remove references to OASIcs series (logo, DOI, ...),
% e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The example in Figure~\ref{fig:factorial} is adapted from the Stainless
documentation \cite{Stainless:documentation} and shows how the verifier is
used. Note how a precondition is specified using \texttt{require} and
a postcondition using \texttt{ensuring}. Our function does not satisfy
the specification. An overflow in the 32-bit integer type leads to a
negative result for the input 17, as Stainless reports in
Figure~\ref{fig:failed}. Changing the type \texttt{Int} to
\texttt{BigInt} will result in a successful verification.

@@

\paragraph{The Pure Scala Fragment.} The Scala fragment supported by
Stainless comprises algebraic data types in the form of abstract
classes, case classes and case objects, objects for grouping classes
and functions, boolean expressions with short-circuit interpretation,
generics with invariant type parameters, pattern matching, local and
anonymous classes and more.  In addition to Pure Scala Stainless also
supports some imperative features, such as while loops and using a
(mutable) variable in a local scope of a function. They turn out not
to be relevant for our current work.

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
can not generate new coins. We call it the \emph{No-Inflation
  Property}. Trying to verify it, we uncover and fix a bug in the
Bitcoin-S library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. So we turn to a simpler property to verify. The
simplest possible property we can think of is the fact that adding
zero satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{Addition-With-Zero Property} and we try
to verify it in Section~3. Here as well we see that a significant part
of the code lies outside of the supported fragment. We rewrite it
until we arrive at code that we successfully verify. In that process
we find and fix a second bug in Bitcoin-S.

@@

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire Bitcoin-S codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
No-Inflation Property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment. We started to rewrite the code in
the supported fragment, but quickly realized that a better approach is
to first verify a simpler property depending on less code and later
come back to the No-Inflation Property with more experience. However,
during the process of trying to rewrite the code, we found a bug in
the \texttt{checkTransaction} function shown in
Figure~\ref{fig:checktrans}.

\paragraph{A Bug in the checkTransaction Function.} Given a
transaction the function returns true if some basic checks succeed,
otherwise false. For example, one of those checks is that both the
list of inputs and list of outputs need to be non-empty.

@@

\begin{figure}
\begin{lstlisting}[style=scala, firstnumber=15]
  val prevOutputs = transaction.inputs.map(_.previousOutput)
  val noDuplicateInputs = 
    prevOutputs.distinct.size == prevOutputs.size
\end{lstlisting}
  \caption{Bug Fix}
  \label{fig:bugfix1}
\end{figure}

Note particularly lines 15-17.  Here, the value
\texttt{prevOutputTxIds} gathers a list of all transaction identifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict:
\texttt{checkTransaction} returns false for valid transactions.

@@

Specifically, we replace lines 15-17 as shown in
Figure~\ref{fig:bugfix1}. We submitted this fix together with a
corresponding unit test to the Bitcoin-S project in a pull request,
which has been merged \cite{BitcoinS:pull435}.

We now turn to the much simpler Addition-With-Zero Property.

\section{The Addition-With-Zero Property}

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in Bitcoin-S is non-trivial. The
reason for that is a peculiarity of consensus code: agreement with the
majority is the only relevant notion of correctness. The most widely
used bitcoin implementation by far is the reference implementation
Bitcoin Core, written in C++. For consensus code, Bitcoin-S thus has
little choice but to be in strict agreement with the reference
implementation. To achieve that, it implements C-like data types in
Scala and then implements functionality using those C-like data
types. For example, the Satoshis class, which represents an amount of
satoshis, is implemented using the class \texttt{Int64} which aims to
represent the C-type \texttt{int64\_t}.

@@

While this is a potentially serious bug, it turns out that
\texttt{checkResult} is only ever called inside a constructor call for
a number type which contains the intended range check, see lines
32-35. The \texttt{checkResult} function thus can, and should, be
removed entirely. The Bitcoin-S developers have acknowledged the bug
and we submitted a pull request to fix it, which has been merged
\cite{BitcoinS:pull565}.

For further development of Bitcoin-S, this raises a question. If the
goal of the \texttt{Int64} type is to emulate \texttt{int64\_t} then
why does it prevent overflows? To achieve strict agreement with
Bitcoin Core, a better approach might be to remove overflow checking
from the data type and to add it in exactly those places where it
happens in Bitcoin Core.

@@

All code changes are \emph{equivalent} in the (admittedly narrow)
sense that if the Addition-With-Zero Property holds for the rewritten
code, then it also holds for the original code.

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those objects into
case objects. That code is equivalent: case objects have various
additional properties (for example, being serializable) but none of
our code depends on the absence of those.

@@

\paragraph{The Specification.} Now that all our code is in the
supported fragment, we can finally write our specification. We add a
postcondition to the +-method of the \texttt{CurrencyUnit}-class
(Figure~\ref{fig:currencyunits}, lines 9-11) resulting in
Figure~\ref{fig:spec}. We successfully verify it with Stainless, as
the output in Figure~\ref{fig:result} shows.

@@

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{result_output}
	\caption{Stainless output for the rewritten code}
  \label{fig:result}
\end{figure}
